Extreme heterogeneity of influenza virus infection in single cells

### 2018-01-12 paper/ResponseToReviews.tex

The strengths of the paper include the description of a novel method for leveraging single-cell RNAseq to quantify heterogeneity and stochasticity in viral gene expression during infection. This is a really important question for the field, and the approaches detailed here will likely by adopted by other groups studying other viruses. The authors do a good job of accounting for many of the pitfalls inherent in the experimental execution and data analysis. The manuscript is well written, and does an especially effective job of visually presenting complex datasets.

The main shortcoming of the paper is that it provides very little in the way of new information. As the authors clearly and honestly point out, most of the findings in this paper are simply confirming observations made in older papers. Additionally, the findings are purely descriptive, and provide little insight into the mechanisms that may give rise to the observations.

We believe that addressing the points below would help ameliorate some of the shortcomings of the paper.

{\color{black}
This is a fair summary of our work.
Thank you for recognizing the power of the approach and our efforts to fully describe pitfalls and effectively present the complex datasets.

@@

    b. Use simple models to illustrate how much variability can be accounted for by simple effects, such as expected Poisson co-infection frequency, or the likelihood of attaining full complement of RNP genes.

{\color{red}
Gini index is something we could definitely calculate. Not clear what we could compare it to, but we can at least calculate it and report it.
We then calculate Gini index under some simple model where everything is due to decreased transcription when missing segments.
Any such model is probably pretty bogus and oversimplified, so we'd want to do this in a way where everything ends up in a heavily caveated supplement that doesn't bog down text.
}

2. While the silent tagging method used to address co-infection is clever and appreciated, the issue is not fully settled. The dismissal of co-infection as a factor influencing cell-to-cell variation in lines 127-9 is based on too few cells to draw any conclusions, and thus needs to be tempered. Also, there is likely to be a significant amount of cryptic co-infection with identical barcode viruses (expected to be similar to that of mixed barcodes, ~10\%) that could influence measured heterogeneity. These points should be made in the text.

{\color{black}
These are very good points.
We have added the following text that clearly draws out the fact that our approach will only identify half the co-infections:
\begin{quote}
\textsl{
In addition, the synonymous viral barcodes only identify co-infections by viruses with different barcodes -- since the barcodes are at roughly equal proportion, we expect to miss about half of the co-infections.
Since we annotate about $\sim$10\% of the infected cells as co-infected by viruses with different barcodes (Figure 4D), we expect another $\sim$10\% of the infected cells to also be co-infected but not annotated as so by our approach.
}
\end{quote}

The reviewers are correct that our statement about the role of co-infection in influencing cell-to-cell variation was too strong.
We have clarified this statement.
We have also added a figure supplement (Figure 4-Figure supplement 3) which uses an independent method (flow cytometry staining of individual viral proteins) to support the fact that even at high infectious dose both low- and high-producing cells remain, although their relative proportions change.
This figure supplement also supports our clarified statement that viral mRNA (or protein) production is not a simple continuous function of infectious dose, but is also affected by other sources of heterogeneity.
The new text reads:
\begin{quote}
\textsl{
Notably, Figure 4E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (Figure 4-Figure supplement 3).
This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (Figure 4-Figure supplement 3).
}
\end{quote}
}

### 2018-01-12 paper/paper.tex
finished text updates about coinfection and viral production

\figsupp[Flow cytometry analysis of viral protein production of cells contingent on infectious dose or coinfection state. \label{figsupp:flowdose}]
{
(Top) Expression of a viral proteins in cells infected at high (MOI 5 as calculated by TCID50 on MDCK-SIAT1 cells) or low (MOI 0.1) initial infectious dose.
Cells were concurrently stained for HA and NS1 10 hours post infection.
HA staining was analyzed in cells positive for NS1 (left), and NS1 staining was analyzed in cells positive for HA (right).
While a higher dose leads to more cells expressing high amounts of viral protein, it does not substantially increase the amount of viral protein in either the low-expressing or high-expressing cells. 
Therefore, higher viral dose does not lead to a large continuous increase in viral protein production among all cells -- rather, it mostly changes the distribution of where cells fall in the highly heterogeneous distribution of viral protein production.
(Bottom) Cells were co-infected with a mix of wild-type virus and virus in which the HA gene was replaced by GFP flanked by the terminal regions of the HA gene segment. 
At 10 hours post-infection, cells were stained for NS1 and HA expression and analyzed by flow cytometry.
Cells could be annotated as infected by one or more infectious particles (wild-type or pseudovirus alone) or at least two infectious particles (coinfection).
Coinfected cells, like cells infected at a higher infectious dose, occupy different positions in the distribution of viral protein production but do not not exhibit a continuous increase in viral protein production.
Replicates shown.
}{\includegraphics[width=\linewidth]{figures/Coinf.jpg}}
\end{figure}

We determined the threshold amount of viral mRNA per cell at which the barcode partitioning clearly resulted from infection rather than leakage (Figure~\ref{fig:viralbarcodes}C), and used this threshold to annotate cells that we were confident were truly infected.
We also annotated as co-infected cells above this threshold that had mRNA from both viral variants.
Figure~\ref{fig:viralbarcodes}D shows the number of cells annotated as infected and co-infected for each sample -- these cells are just a small fraction of the number of cells with any viral read.
These annotation thresholds are conservative, and will miss some true low-level infections, as well as any co-infections with the same viral variant.
However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the loss of some low-level infections in order to avoid false positives.
Because most cells are not infected, we subsampled the uninfected cells to the numbers shown in Figure~\ref{fig:viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.

Strikingly, the extreme variation in the number of viral transcripts per cell remains even after we apply these rigorous criteria for annotating infected cells (Figure~\ref{fig:viralbarcodes}E). 
The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $\approx$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
Notably, co-infected cells do not express dramatically more viral mRNA than other cells (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:flowdose}), suggesting that the initial infectious dose is not the major driver of cell-to-cell variation in our low-MOI infections.

>>>>

\figsupp[Flow cytometry analysis of viral protein production of cells contingent on infectious dose or coinfection state. \label{figsupp:flowdose}]
{
(Top) Expression of viral proteins in cells infected at high (MOI 5 as calculated by TCID50 on MDCK-SIAT1 cells) or low (MOI 0.1) initial infectious dose.
Cells were concurrently stained for HA and NS1 proteins 10 hours post infection.
HA staining was analyzed in cells positive for NS1 (left), and NS1 staining was analyzed in cells positive for HA (right).
While a higher dose leads to more cells expressing high amounts of viral protein, it does not greatly increase the amount of viral protein in either the low-expressing or high-expressing cells. 
Therefore, higher viral dose does not lead to a large continuous increase in viral protein production among all cells -- rather, it mostly changes the proportions of cells that fall in different parts of the highly heterogeneous distribution.
(Bottom) Cells were co-infected with a mix of wild-type virus and virus in which the HA gene was replaced by GFP flanked by the terminal regions of the HA gene segment. 
At 10 hours post-infection, cells were stained for NS1 and HA expression and analyzed by flow cytometry.
Cells could be annotated as infected by one or more infectious particles (wild-type or pseudovirus alone) or at least two infectious particles (coinfection).
Coinfected cells, like cells infected at a higher infectious dose, occupy different positions in the distribution of viral protein production but do not not exhibit a continuous increase in viral protein production.
Replicates shown.
}{\includegraphics[width=\linewidth]{figures/Coinf.jpg}}
\end{figure}

We determined the threshold amount of viral mRNA per cell at which the barcode partitioning clearly resulted from infection rather than leakage (Figure~\ref{fig:viralbarcodes}C), and used this threshold to annotate cells that we were confident were truly infected.
We also annotated as co-infected cells above this threshold that had mRNA from both viral variants.
Figure~\ref{fig:viralbarcodes}D shows the number of cells annotated as infected and co-infected for each sample -- these cells are just a small fraction of the number of cells with any viral read.
These annotation thresholds are conservative, and may miss some true low-level infections.
However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the possible loss of some low-level infections in order to avoid false positives.
In addition, the synonymous viral barcodes only identify co-infections by viruses with different barcodes -- since the barcodes are at roughly equal proportion, we expect to miss about half of the co-infections.
Since we annotate about $\sim$10\% of the infected cells as co-infected by viruses with different barcodes (Figure~\ref{fig:viralbarcodes}D), we expect another $\sim$10\% of the infected cells to also be co-infected but not annotated as so by our approach.
Because most cells are not infected, we subsampled the uninfected cells to the numbers shown in Figure~\ref{fig:viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.

Strikingly, the extreme variation in the number of viral transcripts per cell remains even after we apply these rigorous criteria for annotating infected cells (Figure~\ref{fig:viralbarcodes}E). 
The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $\approx$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
Notably, Figure~\ref{fig:viralbarcodes}E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:flowdose}).
This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:flowdose}).
